Type Safe Redis Queries

### 2017-07-13 TyDe2017/sections/Discussions-orig.tex
Auto stash before merge of "master" and "origin/master"

\section{Type-Level Dictionaries}
\label{sec:type-level-dict}

One of the challenges of statically ensuring type correctness of \Redis{},
which also presents in other stateful languages, is that the type of the value
associated to a key can be altered after updating. To ensure type correctness,
we keep track of the types of all existing keys in a {\em dictionary} ---
conceptually, an associate list, or a list of pairs of keys and \Redis{} types.
For example, the dictionary \ensuremath{[\mskip1.5mu (\text{\tt \char34 A\char34},\Conid{Int}),(\text{\tt \char34 B\char34},\Conid{Char}),(\text{\tt \char34 C\char34},\Conid{Bool})\mskip1.5mu]} represents
a predicate stating that ``the keys \ensuremath{\text{\tt \char34 A\char34}}, \ensuremath{\text{\tt \char34 B\char34}}, and \ensuremath{\text{\tt \char34 C\char34}} are respectively
associated to values of type \ensuremath{\Conid{Int}}, \ensuremath{\Conid{Char}}, and \ensuremath{\Conid{Bool}}.''

The dictionary mixes values (strings such as \ensuremath{\text{\tt \char34 A\char34}}, \ensuremath{\text{\tt \char34 B\char34}}) and types. Further
more, as mentioned in Section~\ref{sec:indexed-monads}, the dictionaries are
to be used parameters to the indexed monad \ensuremath{\Conid{Edis}}. In a dependently typed
programming language (without the so-called ``phase distinction'' ---
separation between types and terms), this would pose no problem. In Haskell
however, the dictionaries, to index a monad, has to be a type as well.

In this section we describe how to construct a type-level dictionary, to be
used with the indexed monad in Section~\ref{sec:indexed-monads}.

\subsection{Datatype Promotion}

Haskell maintains the distinction between values, types, and kinds: values are
categorized by types, and types are categorized by kinds. The kinds are relatively simple: \ensuremath{\mathbin{*}} is the kind of all {\em lifted} types, while type
constructors have kinds such as \ensuremath{\mathbin{*}\to \mathbin{*}}, \ensuremath{\mathbin{*}\to \mathbin{*}\to \mathbin{*}}, etc. Consider the
datatype definitions below:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{data}\;\Conid{Nat}\mathrel{=}\Conid{Zero}\mid \Conid{Suc}\;\Conid{Nat}~~,\qquad\;\mathbf{data}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\mathrel{=}[\mskip1.5mu \mskip1.5mu]\mid \Varid{a}\mathbin{:}[\mskip1.5mu \Varid{a}\mskip1.5mu]~~.{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
The lefthand side is usually seen as having defined a type \ensuremath{\Conid{Nat}\mathbin{::}\mathbin{*}},
and two value constructors \ensuremath{\Conid{Zero}\mathbin{::}\Conid{Nat}} and \ensuremath{\Conid{Suc}\mathbin{::}\Conid{Nat}\to \Conid{Nat}}. The righthand
side is how Haskell lists are understood. The {\em kind} of \ensuremath{[\mskip1.5mu \mathbin{\cdot}\mskip1.5mu]} is \ensuremath{\mathbin{*}\to \mathbin{*}},
since it takes a lifted type \ensuremath{\Varid{a}} to a lifted type \ensuremath{[\mskip1.5mu \Varid{a}\mskip1.5mu]}. The two value constructors respectively have types \ensuremath{[\mskip1.5mu \mskip1.5mu]\mathbin{::}[\mskip1.5mu \Varid{a}\mskip1.5mu]} and \ensuremath{(\mathbin{:})\mathbin{::}\Varid{a}\to [\mskip1.5mu \Varid{a}\mskip1.5mu]\to [\mskip1.5mu \Varid{a}\mskip1.5mu]}, for all type \ensuremath{\Varid{a}}.

The GHC extension \emph{data kinds}~\cite{promotion}, however, automatically
promotes certain ``suitable'' types to kinds.\footnote{It is only informally
described in the GHC manual what types are ``suitable''.} With the extension,
the \ensuremath{\mathbf{data}} definitions above has an alternative reading: \ensuremath{\Conid{Nat}} is a new kind,
\ensuremath{\Conid{Zero}\mathbin{::}\Conid{Nat}} is a type having kind \ensuremath{\Conid{Nat}}, and \ensuremath{\Conid{Suc}\mathbin{::}\Conid{Nat}\to \Conid{Nat}} is a type
constructor, taking a type in kind \ensuremath{\Conid{Nat}} to another type in \ensuremath{\Conid{Nat}}. Whether a
constructor is promoted can often be inferred from the context. To be more specific, prefixing a constructor with a single quote, such as in \ensuremath{\mbox{\textquotesingle}\Conid{Zero}} and
\ensuremath{\mbox{\textquotesingle}\Conid{Suc}}, denotes that it is promoted.

The situation of lists is similar: for all kind \ensuremath{\Varid{k}}, \ensuremath{[\mskip1.5mu \Varid{k}\mskip1.5mu]} is also a kind. For
all kind \ensuremath{\Varid{k}}, \ensuremath{[\mskip1.5mu \mskip1.5mu]\mathbin{::}[\mskip1.5mu \Varid{k}\mskip1.5mu]} is a type. Given a type \ensuremath{\Varid{x}}
of kind \ensuremath{\Varid{k}} and a type \ensuremath{\Varid{xs}} of kind \ensuremath{[\mskip1.5mu \Varid{k}\mskip1.5mu]}, \ensuremath{\Varid{x}\mathbin{:}\Varid{xs}} is again a type of
kind \ensuremath{[\mskip1.5mu \Varid{k}\mskip1.5mu]}. Formally, \ensuremath{(\mathbin{:})\mathbin{::}\Varid{k}\to [\mskip1.5mu \Varid{k}\mskip1.5mu]\to [\mskip1.5mu \Varid{k}\mskip1.5mu]}. For example,
\ensuremath{\Conid{Int}\mathbin{\mbox{\textquotesingle}\!:}(\Conid{Char}\mathbin{\mbox{\textquotesingle}\!:}(\Conid{Bool}\mathbin{\mbox{\textquotesingle}\!:}\mbox{\textquotesingle}[~]))} is a type having kind \ensuremath{[\mskip1.5mu \mathbin{*}\mskip1.5mu]} --- a list of
(lifted) types. The optional quote denotes that the constructors are promoted.
The same list can be denoted by a syntax sugar \ensuremath{\mbox{\textquotesingle}[\Conid{Int},\Conid{Char},\Conid{Bool}]}.

Tuples are also promoted. Thus we may put two types in a pair to form another
type, such as in \ensuremath{\mbox{\textquotesingle}(\Conid{Int},\Conid{Char})}, a type having kind \ensuremath{(\mathbin{*},\mathbin{*})}.

Strings in Haskell are nothing but lists of \ensuremath{\Conid{Char}}s. Regarding promotion,
however, a \ensuremath{\Conid{String}} can be promoted to a type having kind \ensuremath{\Conid{Symbol}}. \ensuremath{\Conid{Symbol}} is
a type without a constructor: \ensuremath{\mathbf{data}\;\Conid{Symbol}},
intended to be used as a promoted kind. In the expression:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\text{\tt \char34 this~is~a~type-level~string~literal\char34}\mathbin{::}\Conid{Symbol}~~,{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
the string on the lefthand side of \ensuremath{(\mathbin{::})} is a type, having kind \ensuremath{\Conid{Symbol}}.

With all of these ingredients, we are ready to build our dictionaries, or
type-level associate lists:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{type}\;\Conid{DictEmpty}\mathrel{=}\mbox{\textquotesingle}[~]~~,{}\<[E]%
\\
\>[B]{}\mathbf{type}\;\Conid{Dict0}\mathrel{=}\mbox{\textquotesingle}[\mbox{\textquotesingle}(\text{\tt \char34 key\char34},\Conid{Bool})]~~,{}\<[E]%
\\
\>[B]{}\mathbf{type}\;\Conid{Dict1}\mathrel{=}\mbox{\textquotesingle}[\mbox{\textquotesingle}(\text{\tt \char34 A\char34},\Conid{Int}),\mbox{\textquotesingle}(\text{\tt \char34 B\char34},\text{\tt \char34 A\char34})]~~.{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
All the entities defined above are types, where \ensuremath{\Conid{Dict0}} and \ensuremath{\Conid{Dict1}}
have kind \ensuremath{[\mskip1.5mu (\Conid{Symbol},\mathbin{*})\mskip1.5mu]}.

\subsection{Type-Level Functions}

Now that we can represent dictionaries as types, the next step is to define
operations on them. A function that inserts an entry to a dictionary, for
example, is a function from a type to a type. While it was shown that it is
possible to simulate type-level functions using Haskell type
classes~\cite{McBride:02:Faking}, in recent versions of GHC, {\em indexed type
families}, or type families for short, are considered a cleaner solution.

For example, compare conjunction \ensuremath{(\mathrel{\wedge})} and its type-level
counterpart \ensuremath{\Conid{And}}:\\
\noindent{\centering %\small
\begin{minipage}[b]{0.35\linewidth}
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{7}{@{}>{\hspre}c<{\hspost}@{}}%
\column{7E}{@{}l@{}}%
\column{11}{@{}>{\hspre}l<{\hspost}@{}}%
\column{17}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}(\mathrel{\wedge})\mathbin{::}\Conid{Bool}\to \Conid{Bool}\to \Conid{Bool}{}\<[E]%
\\
\>[B]{}\Conid{True}{}\<[7]%
\>[7]{}\mathrel{\wedge}{}\<[7E]%
\>[11]{}\Conid{True}{}\<[17]%
\>[17]{}\mathrel{=}\Conid{True}{}\<[E]%
\\
\>[B]{}\Varid{a}{}\<[7]%
\>[7]{}\mathrel{\wedge}{}\<[7E]%
\>[11]{}\Varid{b}{}\<[17]%
\>[17]{}\mathrel{=}\Conid{False}~~,{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
\end{minipage}
\begin{minipage}[b]{0.55\linewidth}
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{10}{@{}>{\hspre}l<{\hspost}@{}}%
\column{15}{@{}>{\hspre}l<{\hspost}@{}}%
\column{21}{@{}>{\hspre}l<{\hspost}@{}}%
\column{27}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{type}\;\Varid{family}\;\Conid{And}\;(\Varid{a}\mathbin{::}\Conid{Bool})\;(\Varid{b}\mathbin{::}\Conid{Bool})\mathbin{::}\Conid{Bool}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\mathbf{where}\;{}\<[10]%
\>[10]{}\Conid{And}\;{}\<[15]%
\>[15]{}\Conid{True}\;{}\<[21]%
\>[21]{}\Conid{True}{}\<[27]%
\>[27]{}\mathrel{=}\Conid{True}{}\<[E]%
\\
\>[10]{}\Conid{And}\;{}\<[15]%
\>[15]{}\Varid{a}\;{}\<[21]%
\>[21]{}\Varid{b}{}\<[27]%
\>[27]{}\mathrel{=}\Conid{False}~~.{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
\end{minipage}
}\\
The lefthand side is a typical definition of \ensuremath{(\mathrel{\wedge})} by pattern matching.
On the righthand side, \ensuremath{\Conid{Bool}} is not a type, but a type lifted to a kind,
while \ensuremath{\Conid{True}} and \ensuremath{\Conid{False}} are types of kind \ensuremath{\Conid{Bool}}. The declaration says
that \ensuremath{\Conid{And}} is a family of types, indexed by two parameters \ensuremath{\Varid{a}} and \ensuremath{\Varid{b}} of
kind \ensuremath{\Conid{Bool}}. The type with index \ensuremath{\Conid{True}} and \ensuremath{\Conid{True}} is \ensuremath{\Conid{True}}, and all
other indices lead to \ensuremath{\Conid{False}}. For our purpose, we can read \ensuremath{\Conid{And}} as a type-level function. Observe how it resembles the term-level \ensuremath{(\mathrel{\wedge})}.

Note that type families in Haskell come in many flavors. Families can be
defined for \ensuremath{\mathbf{data}} and \ensuremath{\mathbf{type}} synonym. They can appear inside type
classes~\cite{tfclass,tfsynonym} or at toplevel. Toplevel type families
can be open~\cite{tfopen} or closed~\cite{tfclosed}. The flavor we chose
is top-level, closed type synonym family, since it allows overlapping
instances, and we need none of the extensibility provided by open type
families. Notice that the instance \ensuremath{\Conid{And}\;\Conid{True}\;\Conid{True}} could be subsumed under
the more general instance, \ensuremath{\Conid{And}\;\Varid{a}\;\Varid{b}}. In a closed type family we may resolve
the overlapping in order, just like how cases overlapping is resolved in term-level functions.

%\subsection{Functions on Type-Level Dictionaries}

We are now able to define operations on type-level dictionaries.
Let's begin with dictionary lookup.
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{32}{@{}>{\hspre}c<{\hspost}@{}}%
\column{32E}{@{}l@{}}%
\column{35}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{type}\;\Varid{family}\;\Conid{Get}\;(\Varid{xs}\mathbin{::}[\mskip1.5mu (\Conid{Symbol},\mathbin{*})\mskip1.5mu])\;(\Varid{s}\mathbin{::}\Conid{Symbol})\mathbin{::}\mathbin{*}\mathbf{where}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Conid{Get}\;(\mbox{\textquotesingle}(\Varid{s},\Varid{x})\mathbin{\mbox{\textquotesingle}\!:}\Varid{xs})\;\Varid{s}{}\<[32]%
\>[32]{}\mathrel{=}{}\<[32E]%
\>[35]{}\Varid{x}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Conid{Get}\;(\mbox{\textquotesingle}(\Varid{t},\Varid{x})\mathbin{\mbox{\textquotesingle}\!:}\Varid{xs})\;\Varid{s}{}\<[32]%
\>[32]{}\mathrel{=}{}\<[32E]%
\>[35]{}\Conid{Get}\;\Varid{xs}\;\Varid{s}~~{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
The type-level function \ensuremath{\Conid{Get}} returns the entry associated with key \ensuremath{\Varid{s}} in the
dictionary \ensuremath{\Varid{xs}}. Notice, in the first case, how type-level equality can be
expressed by unifying type variables with the same name.
Note also that \ensuremath{\Conid{Get}} is a partial function on types:
\ensuremath{\Conid{Get}\;\mbox{\textquotesingle}[\mbox{\textquotesingle}(\text{\tt \char34 A\char34},\Conid{Int})]\;\text{\tt \char34 A\char34}} evaluates to \ensuremath{\Conid{Int}}, but
\ensuremath{\Conid{Get}\;\mbox{\textquotesingle}[\mbox{\textquotesingle}(\text{\tt \char34 A\char34},\Conid{Int})]\;\text{\tt \char34 B\char34}} gets stuck.
... and these types are computed at compile-time. It wouldn't make
much sense for a type checker to crash and throw a ``Non-exhaustive'' error or
be non-terminating.
\todo{So, what exactly happens when we do the second?}

We could make \ensuremath{\Conid{Get}} total, as we would at the term level, with \ensuremath{\Conid{Maybe}}:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{29}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{type}\;\Varid{family}\;\Conid{Get}\;(\Varid{xs}\mathbin{::}[\mskip1.5mu (\Conid{Symbol},\mathbin{*})\mskip1.5mu])\;(\Varid{s}\mathbin{::}\Conid{Symbol})\mathbin{::}\Conid{Maybe}\mathbin{*}\mathbf{where}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Conid{Get}\;\mbox{\textquotesingle}[~]\;{}\<[29]%
\>[29]{}\Varid{s}\mathrel{=}\Conid{Nothing}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Conid{Get}\;(\mbox{\textquotesingle}(\Varid{s},\Varid{x})\mathbin{\mbox{\textquotesingle}\!:}\Varid{xs})\;{}\<[29]%
\>[29]{}\Varid{s}\mathrel{=}\Conid{Just}\;\Varid{x}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Conid{Get}\;(\mbox{\textquotesingle}(\Varid{t},\Varid{x})\mathbin{\mbox{\textquotesingle}\!:}\Varid{xs})\;{}\<[29]%
\>[29]{}\Varid{s}\mathrel{=}\Conid{Get}\;\Varid{xs}\;\Varid{s}~~.{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
%
Some other dictionary-related functions are defined in a similar fashion
in Figure \ref{fig:dict-operations}. The function \ensuremath{\Conid{Set}} either updates an
existing entry or inserts a new entry, \ensuremath{\Conid{Del}} removes an entry matching
a given key, while \ensuremath{\Conid{Member}} checks whether a given key exists in the
dictionary.

\begin{figure}
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{18}{@{}>{\hspre}l<{\hspost}@{}}%
\column{29}{@{}>{\hspre}l<{\hspost}@{}}%
\column{32}{@{}>{\hspre}l<{\hspost}@{}}%
\column{33}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mbox{\onelinecomment  inserts or updates an entry}{}\<[E]%
\\
\>[B]{}\mathbf{type}\;\Varid{family}\;\Conid{Set}\;{}\<[18]%
\>[18]{}(\Varid{xs}\mathbin{::}[\mskip1.5mu (\Conid{Symbol},\mathbin{*})\mskip1.5mu])\;(\Varid{s}\mathbin{::}\Conid{Symbol})\;(\Varid{x}\mathbin{::}\mathbin{*})\mathbin{::}[\mskip1.5mu (\Conid{Symbol},\mathbin{*})\mskip1.5mu]\;\mathbf{where}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Conid{Set}\;\mbox{\textquotesingle}[~]\;{}\<[29]%
\>[29]{}\Varid{s}\;\Varid{x}\mathrel{=}\mbox{\textquotesingle}[\mbox{\textquotesingle}(\Varid{s},\Varid{x})]{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Conid{Set}\;(\mbox{\textquotesingle}(\Varid{s},\Varid{y})\mathbin{\mbox{\textquotesingle}\!:}\Varid{xs})\;{}\<[29]%
\>[29]{}\Varid{s}\;\Varid{x}\mathrel{=}\mbox{\textquotesingle}(\Varid{s},\Varid{x})\mathbin{\mbox{\textquotesingle}\!:}\Varid{xs}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Conid{Set}\;(\mbox{\textquotesingle}(\Varid{t},\Varid{y})\mathbin{\mbox{\textquotesingle}\!:}\Varid{xs})\;{}\<[29]%
\>[29]{}\Varid{s}\;\Varid{x}\mathrel{=}\mbox{\textquotesingle}(\Varid{t},\Varid{y})\mathbin{\mbox{\textquotesingle}\!:}\Conid{Set}\;\Varid{xs}\;\Varid{s}\;\Varid{x}{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\mbox{\onelinecomment  removes an entry}{}\<[E]%
\\
\>[B]{}\mathbf{type}\;\Varid{family}\;\Conid{Del}\;{}\<[18]%
\>[18]{}(\Varid{xs}\mathbin{::}[\mskip1.5mu (\Conid{Symbol},\mathbin{*})\mskip1.5mu])\;(\Varid{s}\mathbin{::}\Conid{Symbol})\mathbin{::}[\mskip1.5mu (\Conid{Symbol},\mathbin{*})\mskip1.5mu]\;\mathbf{where}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Conid{Del}\;\Conid{Nil}\;{}\<[29]%
\>[29]{}\Varid{s}{}\<[32]%
\>[32]{}\mathrel{=}\Conid{Nil}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Conid{Del}\;(\mbox{\textquotesingle}(\Varid{s},\Varid{y})\mathbin{\mbox{\textquotesingle}\!:}\Varid{xs})\;\Varid{s}{}\<[32]%
\>[32]{}\mathrel{=}\Varid{xs}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Conid{Del}\;(\mbox{\textquotesingle}(\Varid{t},\Varid{y})\mathbin{\mbox{\textquotesingle}\!:}\Varid{xs})\;\Varid{s}{}\<[32]%
\>[32]{}\mathrel{=}\mbox{\textquotesingle}(\Varid{t},\Varid{y})\mathbin{\mbox{\textquotesingle}\!:}\Conid{Del}\;\Varid{xs}\;\Varid{s}{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\mbox{\onelinecomment  membership}{}\<[E]%
\\
\>[B]{}\mathbf{type}\;\Varid{family}\;\Conid{Member}\;(\Varid{xs}\mathbin{::}[\mskip1.5mu (\Conid{Symbol},\mathbin{*})\mskip1.5mu])\;(\Varid{s}\mathbin{::}\Conid{Symbol})\mathbin{::}\Conid{Bool}\;\mathbf{where}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Conid{Member}\;\Conid{Nil}\;{}\<[33]%
\>[33]{}\Varid{s}\mathrel{=}\Conid{False}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Conid{Member}\;(\mbox{\textquotesingle}(\Varid{s},\Varid{x})\mathbin{\mbox{\textquotesingle}\!:}\Varid{xs})\;{}\<[33]%
\>[33]{}\Varid{s}\mathrel{=}\Conid{True}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Conid{Member}\;(\mbox{\textquotesingle}(\Varid{t},\Varid{x})\mathbin{\mbox{\textquotesingle}\!:}\Varid{xs})\;{}\<[33]%
\>[33]{}\Varid{s}\mathrel{=}\Conid{Member}\;\Varid{xs}\;\Varid{s}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
\caption{Some operations on type-level dictionaries.}
\label{fig:dict-operations}
\end{figure}

\subsection{Proxies and Singleton Types}

>>>>

\section{Disscussions}
\label{sec:discussions}

% \paragraph{Syntax}
% No one could ignore the glaring shortcoming of the syntax, which occurs mainly
%  in two places: \emph{symbol singletons} and \emph{indexed monad}. We are hoping
%  that these issues could be resolved with future syntactic extensions.

\paragraph{Returning Inferable Types.} \texttt{GET} is yet another command that
is invokable only under a ``well-typed or non-existent'' precondition,
mentioned in Section~\ref{sec:disjunctive-constraints}. It fetches the value of
a key and, if the key does not exist, returns a special value \texttt{nil}. An
error is raised if the value is not a string. In \Edis{} the situation is made
slightly complex, since we parse the string to the type it was supposed to have
encoded from. The \Edis{} version of \ensuremath{\Varid{get}} could be typed:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{6}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{get}{}\<[6]%
\>[6]{}\mathbin{::}(\Conid{KnownSymbol}\;\Varid{k},\Conid{Serialize}\;\Varid{a},\Conid{StringOrNX}\;\Varid{xs}\;\Varid{k}){}\<[E]%
\\
\>[6]{}\Rightarrow \Conid{Proxy}\;\Varid{k}\to \Conid{Edis}\;\Varid{xs}\;\Varid{xs}\;(\Conid{R} \uplus \Conid{Maybe}\;\Varid{a})~~.{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
where \ensuremath{\Conid{StringOrNX}} is defined in Figure~\ref{fig:xxxOrNX}.

The problem with such typing, however, is that \ensuremath{\Varid{a}} cannot be inferred from \ensuremath{\Varid{xs}}
and \ensuremath{\Varid{k}} when \ensuremath{\Varid{k}} does not appear in \ensuremath{\Varid{xs}}. In such situations, to avoid Haskell
complaining about ambiguous type, \ensuremath{\Varid{a}} has to be specified by the caller of
\ensuremath{\Varid{get}}. The user will then be forced to spell out the complete type signature,
only to make \ensuremath{\Varid{a}} explicit.

We think it is more reasonable to enforce that, when \ensuremath{\Varid{get}} is called, the key
should exist in the data store. Thus \ensuremath{\Varid{get}} in \Redis{} has the following type:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{6}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{get}{}\<[6]%
\>[6]{}\mathbin{::}(\Conid{KnownSymbol}\;\Varid{k},\Conid{Serialize}\;\Varid{a},\Conid{Get}\;\Varid{xs}\;\Varid{k}\mathrel{\sim}\Conid{StringOf}\;\Varid{a}){}\<[E]%
\\
\>[6]{}\Rightarrow \Conid{Proxy}\;\Varid{k}\to \Conid{Edis}\;\Varid{xs}\;\Varid{xs}\;(\Conid{R} \uplus \Conid{Maybe}\;\Varid{a})~~,{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
which requires that \ensuremath{(\Varid{k},\Varid{a})} presents in \ensuremath{\Varid{xs}} and thus \ensuremath{\Varid{a}} is inferrable from
\ensuremath{\Varid{xs}} and \ensuremath{\Varid{k}}.

\paragraph{Variable Number of Input/Outputs.} Recall that, in
Section~\ref{sec:proxy-key}, the \Redis{} command \texttt{DEL} takes a variable
number of keys, while our \Edis{} counterpart takes only one. Some \Redis{}
commands take a variable number of arguments as inputs, and some returns
multiple results. Most of them are accurately implemented in \Hedis{}. For
another example of a variable-number input command, the type of \ensuremath{\Varid{sinter}} in
\Hedis{} is shown below:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{\Conid{Hedis}.sinter}\mathbin{::}[\mskip1.5mu \Conid{ByteString}\mskip1.5mu]\to \Conid{Redis}\;(\Conid{R} \uplus [\mskip1.5mu \Conid{ByteString}\mskip1.5mu])~~.{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
It takes a list of keys, values of which are all supposed to be sets, and
computes their intersection (the returned list is the intersected set).

In \Edis{}, for a function to accept a list of keys as input, we have to
specify that all the keys are in the class \ensuremath{\Conid{KnownSymbol}}. It can be done by
defining a datatype, indexed by the keys, serving as a witness that they
are all in \ensuremath{\Conid{KnownSymbol}}. We currently have not implemented such feature and
leave it as a possible future work. For now, we offer commands that take fixed
numbers of inputs. The \Edis{} version of \ensuremath{\Varid{sinter}} has type:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{12}{@{}>{\hspre}l<{\hspost}@{}}%
\column{15}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{sinter}\mathbin{::}{}\<[12]%
\>[12]{}({}\<[15]%
\>[15]{}\Conid{KnownSymbol}\;\Varid{k}_{1},\Conid{KnownSymbol}\;\Varid{k}_{2},\Conid{Serialize}\;\Varid{a},{}\<[E]%
\\
\>[15]{}\Conid{SetOf}\;\Varid{x}\mathrel{\sim}\Conid{Get}\;\Varid{xs}\;\Varid{k}_{1},\Conid{SetOf}\;\Varid{x}\mathrel{\sim}\Conid{Get}\;\Varid{xs}\;\Varid{k}_{2}){}\<[E]%
\\
\>[12]{}\Rightarrow \Conid{Proxy}\;\Varid{k}_{1}\to \Conid{Proxy}\;\Varid{k}_{2}\to \Conid{Edis}\;\Varid{xs}\;\Varid{xs}\;(\Conid{R} \uplus [\mskip1.5mu \Varid{a}\mskip1.5mu])~~.{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

The function \ensuremath{\Varid{hmset}} in \Hedis{} allows one to set the values of many fields
in a hash, while \ensuremath{\Varid{hgetall}} returns all the field-value pairs of a hash. They
have the following types:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{16}{@{}>{\hspre}l<{\hspost}@{}}%
\column{33}{@{}>{\hspre}l<{\hspost}@{}}%
\column{65}{@{}>{\hspre}l<{\hspost}@{}}%
\column{98}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{\Conid{Hedis}.hmset}{}\<[16]%
\>[16]{}\mathbin{::}\Conid{ByteString}{}\<[33]%
\>[33]{}\to [\mskip1.5mu (\Conid{ByteString},\Conid{ByteString})\mskip1.5mu]{}\<[65]%
\>[65]{}\to \Conid{Redis}\;(\Conid{R} \uplus \Conid{Status}){}\<[98]%
\>[98]{}~~,{}\<[E]%
\\
\>[B]{}\Varid{\Conid{Hedis}.hgetall}{}\<[16]%
\>[16]{}\mathbin{::}\Conid{ByteString}\to \Conid{Redis}\;(\Conid{R} \uplus [\mskip1.5mu (\Conid{ByteString},\Conid{ByteString})\mskip1.5mu])~~.{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
Proper implementations of them in \Edis{} should accept or return a
\emph{heterogeneous list}~\cite{hetero} --- a list whose elements can be of
different types. We also leave such functions as a future work.

\paragraph{Not All Safe Redis Programs Can Be Typechecked.}
Enforcing a typing discipline rules out some programs that are likely to be erroneous, and reduces the number of programs that are allowed. Like all type
systems, our type system takes a conservative estimation: there are bound to be
some \Redis{} programs that are not typable in our type system, but do not
actually throw a type error. We demand that elements in our lists must be
of the same type, for example, while a \Redis{} program could store in a list
different types of data, encoded as strings, and still works well.

One innate limitation is that we cannot allow dynamic generation of keys. In
\Hedis{}, the Haskell program is free to generate arbitrary sequence of keys
to be used in the data store, which is in general not possible due to the
static nature of \Edis{}.

\paragraph{Transactions.} Commands in \Redis{} can be wrapped in
\emph{transactions}. \Redis{} offers two promises regarding commands in a
transaction. Firstly, all commands in a transaction are serialized and
executed sequentially, without interruption from another client. Secondly,
either all of the commands or none are processed.

Support of transactions in \Edis{} is a future work.
We expect that there would not be too much difficulty --- in an early
experiment, we have implemented a runtime type checker specifically targeting
\Redis{} transactions, and we believe that the experience should be applicable
to static type checking as well.
